Fail Faster

### 2025-03-22 paper/paper.tex
blah

These strategies differ in their use of generator library combinators 
and in how frequently they sample random values. The type-derived and 
insertion-based BST generators sample frequently but use combinators 
sparingly. The single-pass BST generator also samples frequently 
but uses more combinators, striking a middle ground. 
The type-derived 
STLC generator samples infrequently and makes moderate 
use of combinators, while the custom STLC generator uses 
significantly more combinators than any other strategy and 
also takes the fewest samples. The boolean list generator 
makes moderate use of combinators 
and samples booleans frequently---however, because booleans are 
inexpensive to sample, its overall sampling overhead is lower 
than that of the other strategies.

For each strategy, we compare three treatments: our \bq baseline; 
a \camlname version using \bq's randomness 
library, \texttt{splittable\_random}; and a
\camlname version using \csplitmix. For each treatment, we measure the
time to generate a value---i.e., a BST, a STLC term, or a boolean list---using a random seed, over a 
variety of different sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc. Regardless of how size is encoded, our goal is to show that performance trends scale.}
We run each treatment for 5 seconds---enough time for several thousand generations---and compute the average generation time of all values produced in that interval.

@@

@@

\subsubsection{\scalaname}
In \scalaname, we implement a subset of the \camlname
strategies: the boolean list generator, single-pass BST generator, and
well-typed STLC term generator. ScalaCheck does not support type derivation,
so those strategies are excluded.

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance
gain---purely from staging---over ScalaCheck than \camlname does over \bq. This difference arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}. Each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging eliminates this
overhead.

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking bug-finding speed}
To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
case studies using Etna, a platform for assessing and comparing PBT techniques.
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. For example, in a tree
with an ordering property, a bug injection might consist of inserting a value
into the tree that violates that ordering. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test different aspects of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations on binary search trees).
In particular, BST has 37 tasks, and STLC has 20. Strategies for BST and STLC
are implemented in
\camlname; they are unchanged from their description in \secref{subsection:benchmarking}.\footnote{Etna does
not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
we see no reason our results in this section would not extend to \scalaname.}

@@

The average individual-task speedups for each strategy and benchmark are shown in
\figref{fig:etna}, with trends in bug-finding speed closely mirroring those in performance
from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals
substantial variability, with most tasks clustering near the median and a long tail of outliers
achieving much larger gains. STLC shows a more bimodal distribution of speedups
than the other benchmarks, which we attribute to the heterogenous difficulty of its tasks:
some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We
find that ``easier'' tasks---those that run on the order of milliseconds---regularly achieve speedups in the
range of 4--9x, whereas tasks that run on the order of seconds achieve more moderate speedups of up to 3.5x.
Overall,
these results give strong evidence that \camlname consistently benefits bug-finding speed, sometimes drastically.
\begin{figure}[h]
  \includegraphics[scale=0.3]{speedup.png}
  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
  }
  \label{fig:etna}
\end{figure}
\begin{figure}[h]
  \includegraphics[scale=0.31]{swarm.png}
  \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.}
  \label{fig:swarm}
\end{figure}

>>>>

% These strategies differ in their use of generator library combinators 
% and in how frequently they sample random values. The type-derived and 
% insertion-based BST generators sample frequently but use combinators 
% sparingly. The single-pass BST generator also samples frequently 
% but uses more combinators, striking a middle ground. 
% The type-derived 
% STLC generator samples infrequently and makes moderate 
% use of combinators, while the custom STLC generator uses 
% significantly more combinators than any other strategy. 
% The boolean list generator 
% makes moderate use of combinators 
% and samples booleans frequently---however, because booleans are 
% inexpensive to sample, its overall sampling overhead is lower 
% than that of the other strategies.

For each strategy, we compare three treatments: our \bq baseline; 
a \camlname version using \bq's randomness 
library, \texttt{splittable\_random}; and a
\camlname version using \csplitmix. For each treatment, we measure the
time to generate a value---i.e., a BST, a STLC term, or a boolean list---using a random seed, over a 
variety of different generation sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc. Regardless of how size is encoded, our goal is to show that performance trends scale.}
We run each treatment for 5 seconds---enough time for several thousand generations---and compute the average generation time of all values produced in that interval.

@@

To investigate the relationship between a generator's use of library 
abstractions and the performance benefits it gains from staging, 
we focused on the \texttt{bind} function---a core component of monadic 
generator libraries, frequently invoked both directly and indirectly 
through other combinators. We instrumented \texttt{bind} to count how many 
times it was called during a generator's execution. Our 
hypothesis was that generators which make heavier use of 
library abstractions like \texttt{bind} would benefit more from staging.

We evaluated this hypothesis using four benchmarks with widely varying 
speedups: insertion-based BSTs, single-pass BSTs, boolean lists, and 
well-typed STLC terms. For each generator, we ran 10,000 trials, 
using a fixed generation size of 100, and recorded the average number 
of \texttt{bind} calls. We then compared these usage statistics to the 
observed speedups.

As shown in \figref{fig:binds}, the results reveal a clear linear 
correlation between the number of \texttt{bind} calls and the speedup 
achieved through staging. This suggests that generators that 
rely more heavily on the library's abstractions tend to 
benefit more from staging optimizations. \tr{TALK ABOUT RANDOMNESS HERE.}

\begin{figure}[h]
  \includegraphics[scale=0.4]{bind.png}
  \caption{}
  \label{fig:binds}
\end{figure}

@@

\subsubsection{\scalaname}
In \scalaname, we implement a subset of the \camlname
strategies: the boolean list generator, single-pass BST generator, and
well-typed STLC term generator\footnote{By default, ScalaCheck does not support type derivation,
so those strategies are excluded.}. We did not implement an optimized version of ScalaCheck's
\rand, but our experimental setup was otherwise the same as in \secref{subsubsection:ocaml}.

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance
gain---purely from staging---over ScalaCheck than \camlname does over \bq. In particular,
the single-pass BST strategy is $4.89-6.51\times$ faster, the boolean list generator is $6.86-13.41\times$ faster,
and STLC is $5.43-9.70\times$ faster. This difference arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}: each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging eliminates this
overhead.

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking bug-finding speed}
To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
case studies using Etna, a platform for assessing and comparing PBT techniques.
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. For example, in a tree
with an ordering property, a bug injection might consist of inserting a value
into the tree that violates that ordering. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test different aspects of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations on binary search trees).
In particular, BST has 37 tasks, and STLC has 20. Strategies for BST and STLC
are implemented in
\camlname; they are unchanged from their description in \secref{subsection:benchmarking}. 
Etna does not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
we see no reason our results in this section would not extend to \scalaname.

@@

Our results are summarized in \figref{fig:etna}, which shows the geometric average of 
individual-task speedups for each strategy and benchmark. Trends in bug-finding speed reflect trends in performance
from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals
substantial variability, with most tasks clustering near the median and a long tail of outliers
achieving much larger gains. STLC shows a more bimodal distribution of speedups
than the other benchmarks, which we attribute to the heterogenous difficulty of its tasks:
some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We
find that ``easier'' tasks---those that run on the order of milliseconds---regularly achieve speedups in the
range of $4-9\times$, whereas tasks that run on the order of seconds achieve more moderate speedups of up to $3.5\times$.
Overall,
these results give strong evidence that \camlname consistently benefits bug-finding speed, sometimes drastically.
\begin{figure}[h]
  \includegraphics[scale=0.3]{speedup.png}
  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
  }
  \label{fig:etna}
\end{figure}
\begin{figure}[h]
  \includegraphics[scale=0.31]{swarm.png}
  \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.}
  \label{fig:swarm}
\end{figure}
